---
title: "Analysis of PCA from Technical Replicates from ADSP R5"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# A. R5 Technical Replicate Data

## A.1 Overview

Technical replicates are samples from the same individual processed through independent library preparations and sequencing runs. This document analyzes PCA results from *technical replicates* in the ADSP R5 post-QC dataset to assess technical variation and sequencing quality.

**What this analysis does:**

1.  **Data Loading**: Reads PCA eigenvectors/eigenvalues and covariate data
2.  **Quality Assessment**: Examines variance distribution via scree plots
3.  **Replicate Concordance**: Calculates within-group standard deviations across all PCs
4.  **Technical Issues**: Identifies replicate groups with unexpectedly high variation
5.  **Visualization**: Creates PC scatter plots colored by technical factors

**Key expectations for technical replicates:**

-   **Near-zero variation** in early PCs (biological signal should be identical)
-   **Replicate split** occurs around PC (n_groups - 1) where noise dominates
-   **Technical factors** (PCR-free vs standard, sequencing round) may explain any observed variation

## A.2 Replicate PCA Protocol

We follow the same protocol as for non-technical replicates:

[**Replicate PCA Protocol**]{.underline}

A principal components analysis (PCA) was implemented using replicate samples identified through identity-by-descent analysis. Only replicate groups that include at least one sample from this sequencing round (r5) are included.

-   Select variants (SNVs or INDELs) meeting all of the following criteria:

    -   polymorphic among replicates

    -   no genotype missingness across replicates

    -   minor allele frequency (MAF)\>0.05

    -   QC criteria from fields in \*.summary file: `GLNexusPass`=1 and `CallBad`=0 and `HiDepth`=0

-   LD prune selected variants (r2 threshold of 0.8)

-   Conduct PCA (EIGENSOFT/EIGENSTRAT) to obtain all eigenvectors (PCs) and eigenvalues

## A.3 Data Description

### A.3.1 File Parameters

```{r file_parameters, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide", results='asis'}
# ============================================================================
# FILE PATHS AND PARAMETERS - MODIFY AS NEEDED
# ============================================================================

# ---- INPUT FILES ----
# PCA file format: CSV with first row containing eigenvalues in columns 4+
#                  Subsequent rows contain: ID, Rep, index, PC1, PC2, ..., PCn
pca_file <- "data/TECH_pruned_evec.csv"

# Covariate file format: CSV with columns including:
#   Required: SampleID, Cohort_Site, Study, Round, Sex, Race, Ethnicity
#   Optional: DX_harmonized, Age_harmonized, APOE_reported, PCR.free, Technical_Replicate
covariate_file <- "data/r5_replicate_covariates_TechReps.csv"

# ---- OUTPUT FILES ----
processed_pca_output <- "processed_tech_pca_data.csv"  # Cleaned PCA data with PC column names
replicate_spreads_cache <- "tech_rep_spreads.rds"      # Cache of SD calculations (delete to recalculate)

# ---- ANALYSIS PARAMETERS ----
# PC Region Definitions - MODIFY THESE TO CHANGE ANALYSIS REGIONS
early_pcs_range <- c(1, 5)           # Early PCs region (PC1-PC5)
split_region_range <- c(6, 16)       # Replicate split region (PC6-PC16)
late_pcs_range <- c(17, NA)          # Late PCs region (PC17-end, NA = to last PC)

# Threshold for flagging high SD in early PCs (indicates technical issues)
# For technical replicates, expect SDs < 1e-4 in early PCs
sd_threshold_early_pcs <- 1e-4

# Threshold for flagging high SD in replicate split region
sd_threshold_split_region <- 0.1

# Threshold for flagging high SD in late PCs
sd_threshold_late_pcs <- 0.05

# Percentile cutoff for highlighting top spreads in plots (0.90 = top 10%)
top_percentile_cutoff <- 0.90

# PC range for zoomed scree plot (min and max)
zoom_pc_min <- 6
zoom_pc_max <- 20

# ============================================================================

# Quick reference for threshold values:
# - sd_threshold_early_pcs: Increase if too many false positives (e.g., 1e-3)
# - sd_threshold_split_region: Decrease to be more stringent (e.g., 0.05)
# - sd_threshold_late_pcs: Decrease to be more stringent (e.g., 0.01)
# ============================================================================

# Validate input files exist
if (!file.exists(pca_file)) {
  stop("PCA file not found: ", pca_file)
}
if (!file.exists(covariate_file)) {
  stop("Covariate file not found: ", covariate_file)
}

cat("**Input files validated:**\n\n")
cat("- PCA file:", pca_file, "\n")
cat("- Covariate file:", covariate_file, "\n\n")
```

### A.3.2 Data Integrity Checks

```{r data_checks, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
# Pre-read checks
pca_data_raw <- read.csv(pca_file, header = FALSE, nrows = 5)

cat("**Data Format Checks:**\n\n")
cat("- PCA file columns:", ncol(pca_data_raw), "\n")
cat("- Expected: Column 1 = ID, Column 2 = Rep, Column 3 = index, Columns 4+ = eigenvalues/PCs\n")

# Check for minimum required columns
if (ncol(pca_data_raw) < 4) {
  stop("ERROR: PCA file has fewer than 4 columns. Expected: ID, Rep, index, eigenvalues/PCs")
}

cat("- Format appears correct\n\n")
```

### A.3.3 Replicate Sets

```{r get_pca_data, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
# Read PCA results file
df <- read.csv(
  pca_file,
  header = FALSE,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Extract eigenvalues (row 1, col 4+)
eigenvalues <- as.numeric(df[1, 4:ncol(df)])
headers <- as.character(df[1, ])
ev_df <- df[-1, ]
colnames(ev_df) <- headers

# Rename EV columns to PC1..PCn
num_pcs <- ncol(ev_df) - 3
pc_names <- paste0("PC", seq_len(num_pcs))
colnames(ev_df)[4:ncol(ev_df)] <- pc_names

# Convert PCs to numeric
invisible(ev_df[pc_names] <- lapply(ev_df[pc_names], as.numeric))

# Save processed data
invisible(write.csv(ev_df, processed_pca_output, row.names = FALSE))

# Standardize column names for consistency
if ("Rep" %in% names(ev_df)) {
  names(ev_df)[names(ev_df) == "Rep"] <- "REP"
}
if ("ID" %in% names(ev_df)) {
  # ID column is already correct
} else {
  stop("Expected 'ID' column not found in PCA data")
}

# Verify we have the expected columns
expected_cols <- c("ID", "REP", "index")
missing_cols <- setdiff(expected_cols, names(ev_df))
if (length(missing_cols) > 0) {
  stop("Missing expected columns: ", paste(missing_cols, collapse = ", "))
}

# Count replicate groups
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

n_groups <- length(unique(ev_df$REP))
n_samples <- nrow(ev_df)

cat("**Technical Replicate Summary:**\n\n")
cat(paste0("- **", n_groups, " Replicate Groups**\n"))
cat(paste0("- **", n_samples, " Total Samples**\n"))
cat(paste0("- **", num_pcs, " Principal Components**\n\n"))

cat("\n**Expected Analysis Behavior:**\n\n")
cat("- Technical replicates should show near-zero variation in early PCs\n")
cat("- Replicate split expected around PC", n_groups - 1, "(n_groups - 1)\n")
cat("- SD > ", sd_threshold_early_pcs, " in early PCs indicates potential technical issues\n\n")
```

### A.3.4 Covariates

```{r get_cov_data, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
library(dplyr)

# Read covariate file and filter to samples in PCA
covariates_raw <- read.csv(covariate_file)

# Check for samples in PCA but not in covariates
samples_missing_cov <- setdiff(ev_df$ID, covariates_raw$SampleID)
if (length(samples_missing_cov) > 0) {
  warning("WARNING: ", length(samples_missing_cov), " samples in PCA are missing from covariate file")
  cat("Samples missing covariates:\n")
  print(head(samples_missing_cov, 10))
  if (length(samples_missing_cov) > 10) {
    cat("... and", length(samples_missing_cov) - 10, "more\n")
  }
}

covariates_filtered <- covariates_raw %>%
  filter(SampleID %in% ev_df$ID) %>%
  select(SampleID, Cohort_Site, Study, Round, Sex, DX_harmonized, Age_harmonized,
         Age_baseline, APOE_reported, Race, Ethnicity, Technical_Replicate, PCR.free,
         Sequencing_Center)

# Recode Ethnicity and Race for readability
covariates_filtered <- covariates_filtered %>%
  mutate(
    Ethnicity = recode(Ethnicity, `0` = "NH", `1` = "H"),
    Race = recode(Race, `1` = "AI", `2` = "As", `4` = "Bl", `5` = "W", `6` = "Other")
  )

cat("**Covariate Summary:**\n\n")
cat("- Samples with covariate data:", nrow(covariates_filtered), "\n")
cat("- Cohort/Sites represented:", length(unique(covariates_filtered$Cohort_Site)), "\n")
cat("- Studies represented:", length(unique(covariates_filtered$Study)), "\n")
cat("- Sequencing rounds:", paste(sort(unique(covariates_filtered$Round)), collapse = ", "), "\n")
if("PCR.free" %in% names(covariates_filtered)) {
  pcr_free_summary <- table(covariates_filtered$PCR.free, useNA = "ifany")
  cat("- Library prep types:\n")
  for (type in names(pcr_free_summary)) {
    if (!is.na(type)) {
      cat("    - ", type, ": ", pcr_free_summary[type], "\n", sep = "")
    } else {
      cat("    - Missing: ", pcr_free_summary[type], "\n", sep = "")
    }
  }
}
if("Sequencing_Center" %in% names(covariates_filtered)) {
  seq_center_summary <- table(covariates_filtered$Sequencing_Center, useNA = "ifany")
  cat("- Sequencing centers:\n")
  for (center in names(seq_center_summary)) {
    if (!is.na(center)) {
      cat("    - ", center, ": ", seq_center_summary[center], "\n", sep = "")
    } else {
      cat("    - Missing: ", seq_center_summary[center], "\n", sep = "")
    }
  }
}
cat("\n")
```

### A.3.5 Replicate Group Details

**Detailed Replicate Group Information:** This section shows the technical covariates for each replicate group. Technical replicates from the same individual have been processed at different sequencing centers, in different rounds, or with different library preparation methods (PCR-free vs PCR-amplified).

-   TECH1, TECH2 and TECH3 are the first replicate groups that were used in the initial ADSP-Discovery sequencing studies. They include replicate samples both PCR-free and PCR-amplified preps in r1 and they were carried forward throughout subsequent sequencing rounds.
  -   The three unique samples are from the same family (TECH1 = mother, TECH2 and TECH3 = full sibling offspring of TECH1)
  -   TECH3 is missing the replicate sample in the LASI-DAD study.
-   TECH4, TECH5 and TECH6 were introduced in the ADSP-Extension studies and have dropped the three PCR-Amplified samples from r1. They were also carried forward in subsequent sequencing rounds.
  -   TECH5 is missing the replicate samples from LASI-DAD and GARD1.
- Based on reported NIH Race and Ethnicity categories:
  - TECH1, TECH2, TECH3 and TECH4 are Hispanic
  - TECH5 is Non-Hispanic white
  - TECH6 is Non-Hispanic black

```{r replicate_details, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
# Merge replicate data with covariates
rep_details <- ev_df %>%
  left_join(covariates_filtered, by = c("ID" = "SampleID")) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center)

# Extract the ending part of sample IDs (after last underscore or dash)
rep_details <- rep_details %>%
  mutate(ID_ending = sub(".*[_-]", "", ID))

# Get unique replicate groups
unique_reps <- sort(unique(rep_details$REP))


# Loop through each replicate group
for (rep_id in unique_reps) {
  rep_data <- rep_details %>%
    filter(REP == rep_id) %>%
    arrange(Round, Study, Sequencing_Center, ID_ending)

  cat("---\n\n")
  cat("**", rep_id, "** (N = ", nrow(rep_data), ")\n\n", sep = "")

  # Show the table
  display_table <- rep_data %>%
    select(ID, Cohort_Site, Study, Round, PCR.free, Sequencing_Center)

  print(knitr::kable(display_table, row.names = FALSE))

}
```

# B. Scree Plots

For technical replicates, most variance should be explained by differences between the replicate groups, which come from different individuals. We expect minimal variance within each replicate group since these are samples from the same individual. Within-group variance should reflect technical differences in library prep and/or differences between the sequencing centers.

```{r scree_plot, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
library(ggplot2)

scree_data <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalue = eigenvalues
)

var_explained <- (eigenvalues / sum(eigenvalues)) * 100
n_pcs_gt1 <- sum(eigenvalues > 1)

# Full scree plot
ggplot(scree_data, aes(x = Component, y = Eigenvalue)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "orange") +
  geom_vline(xintercept = n_samples - 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = n_groups - 1, linetype = "dashed", color = "blue") +
  annotate("text", x = n_samples - 1, y = max(scree_data$Eigenvalue) * 0.85,
           label = paste(n_samples - 1, "= (Samples - 1)"), color = "red", angle = 90, vjust = -0.5) +
  annotate("text", x = n_groups - 1, y = max(scree_data$Eigenvalue) * 0.85,
           label = paste(n_groups - 1, "= (Groups - 1)"), color = "blue", angle = 90, vjust = -0.5) +
  scale_x_continuous(breaks = function(x) {
    breaks <- pretty(x, n = 10)
    breaks <- unique(floor(breaks))
    # Ensure first and last values are included
    if (!1 %in% breaks) breaks <- c(1, breaks)
    if (!length(eigenvalues) %in% breaks) breaks <- c(breaks, length(eigenvalues))
    sort(unique(breaks))
  }) +
  labs(title = "Scree Plot - Technical Replicates", x = "PC Number", y = "Eigenvalue") +
  theme_minimal()

# Zoomed scree plot
zoom_range <- zoom_pc_min:min(zoom_pc_max, length(eigenvalues))
zoom_max_eigenvalue <- max(eigenvalues[zoom_range])

ggplot(scree_data, aes(x = Component, y = Eigenvalue)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "orange") +
  scale_x_continuous(breaks = function(x) {
    breaks <- pretty(x, n = 10)
    breaks <- unique(floor(breaks))
    # Ensure first and last values of zoom range are included
    if (!zoom_pc_min %in% breaks) breaks <- c(zoom_pc_min, breaks)
    if (!min(zoom_pc_max, length(eigenvalues)) %in% breaks) breaks <- c(breaks, min(zoom_pc_max, length(eigenvalues)))
    sort(unique(breaks))
  }) +
  coord_cartesian(xlim = c(zoom_pc_min, min(zoom_pc_max, length(eigenvalues))),
                  ylim = c(0, zoom_max_eigenvalue * 1.05)) +
  labs(title = paste0("Scree Plot (Zoomed PC", zoom_pc_min, "-PC", min(zoom_pc_max, length(eigenvalues)), ")"),
       x = "PC Number", y = "Eigenvalue") +
  theme_minimal()
```

**Interpretation:** The scree plots show the distribution of variance across principal components. The first 5 PCs explain `round(sum(var_explained[1:5]), 2)`% of variance, which is what we expect if between-group differences dominate. We see very low variance in PCs beyond PC5, consistent with low technical variation.

# C. Replicate Splits

Looking directly at within-group variation of each PC tells us more about how important technical variation is in explaining the overall variation. We expect the standrad deviation (SD) to be very close to zero across PCs until PC6, at which point the replicates should begin to separate in the PC space.

```{r setup_replicate_functions, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide", results='asis'}
library(tidyr)
library(dplyr)
library(purrr)
library(kableExtra)
library(tibble)
library(RColorBrewer)

calculate_replicate_spread <- function(data_df, rep_col, pc_cols, cache_file = NULL) {

  if (!is.null(cache_file) && file.exists(cache_file)) {
    message("Loading cached replicate spreads from ", cache_file)
    return(readRDS(cache_file))
  }

  # Validate inputs
  if (!rep_col %in% names(data_df)) {
    stop("Column '", rep_col, "' not found in data. Available columns: ", paste(names(data_df), collapse = ", "))
  }

  spread_list <- list()
  unique_reps <- unique(data_df[[rep_col]])

  for (rep in unique_reps) {
    rep_data <- data_df[data_df[[rep_col]] == rep, ]

    if (nrow(rep_data) < 2) next

    n_samples <- nrow(rep_data)

    spreads <- data.frame(
      Replicate = rep,
      N_samples = n_samples
    )

    for (pc in pc_cols) {
      pc_values <- rep_data[[pc]]
      spreads[[paste0(pc, "_sd")]] <- sd(pc_values)
    }

    spread_list[[length(spread_list) + 1]] <- spreads
  }

  # Check if we found any replicate groups
  if (length(spread_list) == 0) {
    stop("No replicate groups with 2+ samples found. Check that '", rep_col, "' column contains replicate group IDs.")
  }

  all_spreads <- do.call(rbind, spread_list)

  if (!is.null(cache_file)) {
    message("Saving replicate spreads to ", cache_file)
    saveRDS(all_spreads, cache_file)
  }

  return(all_spreads)
}

plot_pc_range <- function(data_df, pc_min, pc_max, top_flag = NULL, sd_threshold = 1e-4) {

  data_sub <- data_df %>%
    filter(PC_num >= pc_min, PC_num <= pc_max)

  pc_levels <- data_sub %>%
    distinct(PC, PC_num) %>%
    arrange(PC_num) %>%
    pull(PC)

  data_sub <- data_sub %>%
    mutate(PC = factor(PC, levels = pc_levels))

  # Get unique replicates and assign colors (matching Section D PCA plots)
  unique_reps <- sort(unique(data_sub$Replicate))
  n_reps <- length(unique_reps)

  # Use more vibrant, bold colors (same as Section D)
  if (n_reps <= 8) {
    rep_colors <- RColorBrewer::brewer.pal(max(3, n_reps), "Dark2")[1:n_reps]
  } else if (n_reps <= 12) {
    rep_colors <- RColorBrewer::brewer.pal(n_reps, "Paired")
  } else {
    # For many replicates, use rainbow with higher saturation
    rep_colors <- rainbow(n_reps, s = 0.8, v = 0.8)
  }
  names(rep_colors) <- unique_reps

  p <- ggplot(data_sub, aes(x = PC, y = SD))

  if (!is.null(top_flag)) {
    p <- p + geom_point(
      data = data_sub %>% filter(!.data[[top_flag]]),
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.4,
      size = 1.2
    )
    p <- p + geom_point(
      data = data_sub %>% filter(.data[[top_flag]]),
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.8,
      size = 1.5
    )
  } else {
    p <- p + geom_point(
      data = data_sub,
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.6,
      size = 1.2
    )
  }

  theme_settings <- if ((pc_max - pc_min) > 50) {
    theme(axis.text.x = element_blank())
  } else {
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
  }

  p <- p +
    labs(
      title = paste0("Technical Replicate SDs: PC", pc_min, "–PC", pc_max),
      x = "Principal Component",
      y = "Standard Deviation",
      color = NULL
    ) +
    scale_color_manual(values = rep_colors) +
    scale_y_continuous(limits = c(0, NA)) +
    theme_minimal() +
    theme_settings

  return(p)
}

print_high_sd_reps <- function(data_df, plot_data, pc_min, pc_max, sd_threshold = 0.1) {

  data_sub <- data_df %>%
    filter(PC_num >= pc_min, PC_num <= pc_max, SD > sd_threshold) %>%
    arrange(desc(SD))

  if (nrow(data_sub) == 0) {
    cat("\n\nNo replicates with SD >", sd_threshold, "in PC", pc_min, "-", pc_max, "\n\n")
    return(invisible(NULL))
  }

  joined_data <- data_sub %>%
    left_join(plot_data, by = c("Replicate" = "REP"))

  all_pc_colnames <- names(joined_data) %>% stringr::str_subset("^PC\\d+$")

  enriched <- joined_data %>%
    mutate(
      Values = purrr::map2(
        .x = seq_len(nrow(joined_data)),
        .y = PC_num,
        .f = function(row_index, pc_num_value) {
          target_col_name <- paste0("PC", pc_num_value)
          return(joined_data[[target_col_name]][row_index])
        }
      )
    )

  summary_table <- enriched %>%
    arrange(PC_num, Replicate) %>%
    transmute(
      PC,
      ID,
      Replicate,
      Cohort_Site,
      Round,
      PCR.free,
      Sequencing_Center,
      SD,
      Values
    )

  cat("\n\n**Technical Replicates with SD >", sd_threshold, "for PC", pc_min, "-", pc_max, ":**\n\n")

  # Group by replicate and process each group
  unique_reps <- unique(summary_table$Replicate)

  for (i in seq_along(unique_reps)) {
    rep_id <- unique_reps[i]
    rep_data <- summary_table %>% filter(Replicate == rep_id)

    # Add bold separator before each new replicate group (except first)
    if (i > 1) {
      cat("\n---\n\n")
    }

    # For each row in this replicate group, sort the Values column and calculate sort key
    for (j in 1:nrow(rep_data)) {
      values <- unlist(rep_data$Values[[j]])

      # Sort values from smallest to largest
      sorted_values <- sort(values)

      # Store the minimum value as a sort key
      rep_data$sort_key[j] <- min(sorted_values)

      # Format sorted values
      formatted_vals <- sprintf("%.4f", sorted_values)

      rep_data$Values_formatted[[j]] <- paste(formatted_vals, collapse = ", ")
    }

    # Sort the entire rep_data by PC then by sort_key (minimum value)
    rep_data <- rep_data %>% arrange(PC, sort_key)

    # Get unique PCs for this replicate, sorted numerically
    unique_pcs <- unique(rep_data$PC)
    # Extract PC numbers and sort numerically
    pc_numbers <- as.numeric(gsub("PC", "", unique_pcs))
    unique_pcs <- unique_pcs[order(pc_numbers)]

    # Start collapsible section for this replicate
    cat("<details>\n")
    cat("<summary><strong>", rep_id, "</strong> (click to expand)</summary>\n\n", sep = "")

    # Print each PC separately with separators
    for (pc_idx in seq_along(unique_pcs)) {
      pc_name <- unique_pcs[pc_idx]
      pc_data <- rep_data %>% filter(PC == pc_name)

      # Add separator between PCs (except first)
      if (pc_idx > 1) {
        cat("\n---\n\n")
      }

      # Create display table
      display_table <- pc_data %>%
        select(PC, ID, Cohort_Site, Round, PCR.free, Sequencing_Center, SD, Values_formatted) %>%
        rename(Values = Values_formatted)

      print(knitr::kable(display_table, row.names = FALSE, digits = 5, escape = FALSE))
      cat("\n")
    }

    # End collapsible section
    cat("</details>\n\n")
  }

  return(invisible(summary_table))
}

find_variable_replicates <- function(replicate_spreads, threshold = 0) {
  
  sd_cols <- grep("_sd$", names(replicate_spreads), value = TRUE)
  
  variable_reps <- replicate_spreads %>%
    mutate(max_sd = pmax(!!!syms(sd_cols))) %>%
    filter(max_sd > threshold) %>%
    select(Replicate, N_samples, max_sd) %>%
    arrange(desc(max_sd))
  
  cat("\n\n**Replicate groups with SD >", threshold, ":**\n\n")
  cat("Total:", nrow(variable_reps), "out of", nrow(replicate_spreads), "groups\n\n")
  
  print(knitr::kable(variable_reps, row.names = FALSE, digits = 4))
  
  return(invisible(variable_reps))
}

# Prepare plot_data
collapse_rare_levels <- function(column, threshold = 75) {
  level_counts <- table(column)
  rare_levels <- names(level_counts[level_counts <= threshold])

  collapsed <- as.character(column)
  collapsed[collapsed %in% rare_levels] <- "Other"
  collapsed <- as.factor(collapsed)

  return(collapsed)
}

covariates_filtered$Cohort_Site_collapsed <- collapse_rare_levels(covariates_filtered$Cohort_Site)

plot_data <- merge(ev_df, covariates_filtered, by.x = "ID", by.y = "SampleID")

plot_data <- plot_data %>%
  mutate(across(c(Cohort_Site, Study, Round, Race, Ethnicity, Sex, DX_harmonized,
                  Cohort_Site_collapsed, PCR.free, Sequencing_Center), as.factor))

plot_data$Ethnicity <- factor(plot_data$Ethnicity, levels = c("NH", "H", "Other"))
plot_data$Race <- factor(plot_data$Race, levels = c("W", "Bl", "As", "AI", "Other"))

# Ensure PCR.free has proper factor levels
if("PCR.free" %in% names(plot_data)) {
  plot_data$PCR.free <- factor(plot_data$PCR.free,
                                levels = c("PCRFree", "PCRAmplified"),
                                labels = c("PCR-free", "PCR-amplified"))
}
```

## C.1 Overall Visualization

```{r rep_splits_overall, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}

pc_cols <- paste0("PC", 1:num_pcs) 

replicate_spreads <- calculate_replicate_spread(
  plot_data,
  rep_col = "REP",
  pc_cols = pc_cols,
  cache_file = replicate_spreads_cache
)

spreads_long <- replicate_spreads %>%
  select(Replicate, N_samples, ends_with("_sd")) %>%
  pivot_longer(cols = ends_with("_sd"),
               names_to = "PC",
               values_to = "SD") %>%
  mutate(PC = gsub("_sd", "", PC),
         PC_num = as.numeric(gsub("PC", "", PC)))

spreads_long <- spreads_long %>%
  group_by(PC) %>%
  mutate(pcut = quantile(SD, top_percentile_cutoff, na.rm = TRUE),
         Top = SD > pcut) %>%
  ungroup()

plot_pc_range(spreads_long, pc_min = 1, pc_max = num_pcs)

variable_reps <- find_variable_replicates(replicate_spreads, threshold = 0)
```

## C.2 Early PCs (Technical Quality)

Early PCs should show near-zero within-group variation in the absence of substantial technical issues.

```{r rep_splits_early, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
plot_pc_range(spreads_long, pc_min = early_pcs_range[1], pc_max = early_pcs_range[2])
print_high_sd_reps(spreads_long, plot_data, pc_min = early_pcs_range[1], pc_max = early_pcs_range[2],
                   sd_threshold = sd_threshold_early_pcs)
```

## C.3 Replicate Split Region

The replicate-split region includes PC6-PC16. Each PC in the region shows a single replicate group with elevated SDs, indicative of the PC separating samples within the group. All groups appear twice in the region except for TECH5, the smallest group.

```{r rep_splits_middle, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
plot_pc_range(spreads_long, pc_min = split_region_range[1], pc_max = split_region_range[2])
print_high_sd_reps(spreads_long, plot_data, pc_min = split_region_range[1], pc_max = split_region_range[2],
                   sd_threshold = sd_threshold_split_region)
```

## C.4 Late PCs

After PC16, within-group variation decreases substantially.

```{r rep_splits_late, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
late_pc_end <- if (is.na(late_pcs_range[2])) num_pcs else late_pcs_range[2]
plot_pc_range(spreads_long, pc_min = late_pcs_range[1], pc_max = late_pc_end)
print_high_sd_reps(spreads_long, plot_data, pc_min = late_pcs_range[1], pc_max = late_pc_end,
                   sd_threshold = sd_threshold_late_pcs)
```


**Interpretation:** This section identifies technical replicates with high standard deviations across different PC ranges, indicating potential technical issues. Replicates showing very low SDs in early PCs (C.2), suggesting biological differences contribute the most to differences. The replicate split region (C.3) shows where technical variation begins to dominate, occuring after PC`r n_groups - 1`.

# D. Eigenvector Plots

Plotting eigenvectors allows us to visualize how samples cluster based on their PC coordinates.

-   For PC1- PC5, we expect to see replicate groups separate but samples within a group should cluster tightly together.

-   PC6- PC16 show which samples within groups separate. Examining differences in technical covariates for these samples will show which covariates most influence variation.

```{r setup_plotting_functions, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide"}
library(ggplot2)
library(kableExtra)

plot_pcs <- function(df, pc_x, pc_y, color_by, shape_by = NULL,
                     color_title = NULL, shape_title = NULL,
                     var_explained = NULL, clusters = NULL) {

  if (is.null(color_title)) color_title <- color_by
  if (is.null(shape_title)) shape_title <- shape_by

  x_label <- pc_x
  y_label <- pc_y
  if (!is.null(var_explained)) {
    pc_x_num <- as.numeric(gsub("PC", "", pc_x))
    pc_y_num <- as.numeric(gsub("PC", "", pc_y))

    # Use 3 decimal places for PC6-16, 2 for others
    x_digits <- if (pc_x_num >= 6 && pc_x_num <= 16) 3 else 2
    y_digits <- if (pc_y_num >= 6 && pc_y_num <= 16) 3 else 2

    x_label <- paste0(pc_x, " (", round(var_explained[pc_x_num], x_digits), "%)")
    y_label <- paste0(pc_y, " (", round(var_explained[pc_y_num], y_digits), "%)")
  }

  plot_data <- df
  if (!is.null(clusters)) {
    plot_data$cluster <- "Unclustered"
    cluster_names <- names(clusters)
    if (is.null(cluster_names)) {
      cluster_names <- paste0("Cluster", seq_along(clusters))
    }

    for (i in seq_along(clusters)) {
      x_range <- clusters[[i]]$x_range
      y_range <- clusters[[i]]$y_range

      in_cluster <- plot_data[[pc_x]] >= x_range[1] &
                    plot_data[[pc_x]] <= x_range[2] &
                    plot_data[[pc_y]] >= y_range[1] &
                    plot_data[[pc_y]] <= y_range[2]

      plot_data$cluster[in_cluster] <- cluster_names[i]
    }
    plot_data$cluster <- factor(plot_data$cluster,
                                levels = c("Unclustered", cluster_names))
    color_by <- "cluster"
    color_title <- "Cluster"
  }

  p <- ggplot(plot_data, aes(x = .data[[pc_x]], y = .data[[pc_y]],
                      color = .data[[color_by]]))

  if (!is.null(shape_by)) {
    p <- p + geom_point(aes(shape = .data[[shape_by]]), size = 3, stroke = 1, alpha = 0.8)
    p <- p + scale_shape_manual(values = c(1, 2, 0, 5, 6), na.value = 4)
  } else {
    p <- p + geom_point(size = 3, stroke = 1, shape = 16, alpha = 0.8)
  }

  # Use custom colors for REP column (replicate groups)
  if (color_by == "REP") {
    unique_reps <- sort(unique(df[[color_by]]))
    n_reps <- length(unique_reps)

    # Use more vibrant, bold colors
    if (n_reps <= 8) {
      rep_colors <- RColorBrewer::brewer.pal(max(3, n_reps), "Dark2")[1:n_reps]
    } else if (n_reps <= 12) {
      rep_colors <- RColorBrewer::brewer.pal(n_reps, "Paired")
    } else {
      # For many replicates, use rainbow with higher saturation
      rep_colors <- rainbow(n_reps, s = 0.8, v = 0.8)
    }
    names(rep_colors) <- unique_reps

    p <- p + scale_color_manual(values = rep_colors)
  }

  if (!is.null(clusters)) {
    p <- p + scale_color_manual(
      values = c("Unclustered" = "grey80",
                 setNames(scales::hue_pal()(length(cluster_names)), cluster_names)),
      breaks = cluster_names
    )
  }

  p <- p +
    labs(x = x_label, y = y_label, color = color_title, shape = shape_title) +
    theme_minimal()

  return(p)
}

# Function to print points in specified ranges or extremes
print_points <- function(df, pc_x, pc_y, x_range = NULL, y_range = NULL, n = 5) {

  if (is.null(x_range) && is.null(y_range)) {
    x_high_idx <- order(df[[pc_x]], decreasing = TRUE)[1:n]
    x_low_idx <- order(df[[pc_x]], decreasing = FALSE)[1:n]
    y_high_idx <- order(df[[pc_y]], decreasing = TRUE)[1:n]
    y_low_idx <- order(df[[pc_y]], decreasing = FALSE)[1:n]

    cat("**Highest", n, "on", pc_x, ":**\n\n")
    print(knitr::kable(df[x_high_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                        "Sequencing_Center", pc_x, pc_y)] %>%
                        arrange(desc(.data[[pc_x]]), desc(.data[[pc_y]])),
                      row.names = FALSE))

    cat("\n**Lowest", n, "on", pc_x, ":**\n\n")
    print(knitr::kable(df[x_low_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                       "Sequencing_Center", pc_x, pc_y)] %>%
                        arrange(.data[[pc_x]], desc(.data[[pc_y]])),
                      row.names = FALSE))

    cat("\n**Highest", n, "on", pc_y, ":**\n\n")
    print(knitr::kable(df[y_high_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                        "Sequencing_Center", pc_x, pc_y)] %>%
                        arrange(desc(.data[[pc_y]]), desc(.data[[pc_x]])),
                      row.names = FALSE))

    cat("\n**Lowest", n, "on", pc_y, ":**\n\n")
    print(knitr::kable(df[y_low_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                       "Sequencing_Center", pc_x, pc_y)] %>%
                        arrange(.data[[pc_y]], desc(.data[[pc_x]])),
                      row.names = FALSE))

    extreme_idx <- unique(c(x_high_idx, x_low_idx, y_high_idx, y_low_idx))
    extreme_points <- df[extreme_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                        "Sequencing_Center", pc_x, pc_y)]

    return(invisible(extreme_points))

  } else {
    x_in_range <- rep(TRUE, nrow(df))
    y_in_range <- rep(TRUE, nrow(df))

    if (!is.null(x_range)) {
      x_in_range <- df[[pc_x]] >= x_range[1] & df[[pc_x]] <= x_range[2]
    }

    if (!is.null(y_range)) {
      y_in_range <- df[[pc_y]] >= y_range[1] & df[[pc_y]] <= y_range[2]
    }

    in_range_idx <- which(x_in_range & y_in_range)

    points_in_range <- df[in_range_idx, c("ID", "REP", "Cohort_Site", "Round", "PCR.free",
                                           "Sequencing_Center", pc_x, pc_y)] %>%
      arrange(.data[[pc_x]], .data[[pc_y]])

    if (nrow(points_in_range) > 0) {
      print(knitr::kable(points_in_range, row.names = FALSE))
    } else {
      cat("No points found in the specified range.\n\n")
    }

    return(invisible(points_in_range))
  }
}
```

## D.1 PC1 vs PC2

PC1 splits the samples into two sets: TECH1, TECH2, TECH3 vs. TECH4, TECH5, TECH6. These are sets with and without the PCR-amplified samples, respectively.

```{r pc_plots_12, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC1", "PC2", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.2 PC2 vs PC3

PC2 and PC3 separate TECH4, TECH5 and TECH6 from each other.

```{r pc_plots_23, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC2", "PC3", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.3 PC3 vs PC4

```{r pc_plots_34, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC3", "PC4", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.4 PC4 vs PC5

PC4 and PC5 separate TECH1, TECH2 and TECH3 from each other. Specifically PC4 separates the mother, TECH1, from her two offspring, and PC5 separates her two full-sibling offspring, TECH2 from TECH3.

```{r pc_plots_45, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC4", "PC5", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.6 Replicate Split Region (PC6-PC16)

### D.6.1 PC6 vs PC7

PC6 separates TECH6: WASHU/PCR-free/ADSP-Extension-Fam sample from others.

PC7 separates TECH4: MEDGENOME/PCR-free/LASI-DAD sample from others.

```{r pc_plots_67_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC6", "PC7", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC6
pc6_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC6_sd = sd(PC6),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC6_sd))

# Find replicate with largest spread on PC7
pc7_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC7_sd = sd(PC7),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC7_sd))

top_pc6_rep <- pc6_spreads$REP[1]
top_pc7_rep <- pc7_spreads$REP[1]

cat("**Largest spread on PC6:**", top_pc6_rep, "(SD =", sprintf("%.5f", pc6_spreads$PC6_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc6_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC6, PC7) %>%
  arrange(PC6)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC7:**", top_pc7_rep, "(SD =", sprintf("%.5f", pc7_spreads$PC7_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc7_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC6, PC7) %>%
  arrange(PC7)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

### D.6.2 PC8 vs PC9

PC8 separates TECH5: WASHU/PCR-free/ADSP-Extension-Fam sample from others.

PC9 separates TECH4: WASHU/PCR-free/ADSP-Extension-Fam sample from others.

```{r pc_plots_89_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC8", "PC9", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC8
pc8_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC8_sd = sd(PC8),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC8_sd))

# Find replicate with largest spread on PC9
pc9_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC9_sd = sd(PC9),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC9_sd))

top_pc8_rep <- pc8_spreads$REP[1]
top_pc9_rep <- pc9_spreads$REP[1]

cat("**Largest spread on PC8:**", top_pc8_rep, "(SD =", sprintf("%.5f", pc8_spreads$PC8_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc8_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC8, PC9) %>%
  arrange(PC8)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC9:**", top_pc9_rep, "(SD =", sprintf("%.5f", pc9_spreads$PC9_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc9_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC8, PC9) %>%
  arrange(PC9)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

### D.6.3 PC10 vs PC11

PC10 separates TECH3: WASHU/PCR-free/ADSP-Extension-Fam sample from others.

PC11 separates TECH2: WASHU/PCR-free/ADSP-Extension-Fam sample followed by WASHU/PCR-free/ADSP-Extension-CC in a gradient from others.

```{r pc_plots_1011_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC10", "PC11", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC10
pc10_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC10_sd = sd(PC10),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC10_sd))

# Find replicate with largest spread on PC11
pc11_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC11_sd = sd(PC11),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC11_sd))

top_pc10_rep <- pc10_spreads$REP[1]
top_pc11_rep <- pc11_spreads$REP[1]

cat("**Largest spread on PC10:**", top_pc10_rep, "(SD =", sprintf("%.5f", pc10_spreads$PC10_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc10_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC10, PC11) %>%
  arrange(PC10)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC11:**", top_pc11_rep, "(SD =", sprintf("%.5f", pc11_spreads$PC11_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc11_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC10, PC11) %>%
  arrange(PC11)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

### D.6.4 PC12 vs PC13

PC12 separates TECH3: WASHU/PCR-free/ADSP-Extension-CC sample from others.

PC13 separates TECH1: WASHU/PCR-free/ADSP-Extension-Fam sample from others.

```{r pc_plots_1213_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC12", "PC13", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC12
pc12_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC12_sd = sd(PC12),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC12_sd))

# Find replicate with largest spread on PC13
pc13_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC13_sd = sd(PC13),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC13_sd))

top_pc12_rep <- pc12_spreads$REP[1]
top_pc13_rep <- pc13_spreads$REP[1]

cat("**Largest spread on PC12:**", top_pc12_rep, "(SD =", sprintf("%.5f", pc12_spreads$PC12_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc12_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC12, PC13) %>%
  arrange(PC12)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC13:**", top_pc13_rep, "(SD =", sprintf("%.5f", pc13_spreads$PC13_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc13_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC12, PC13) %>%
  arrange(PC13)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

### D.6.5 PC14 vs PC15

PC14 separates TECH2: WASHU/PCR-free/ADSP-Extension-CC sample from WASHU/PCR-free/ADSP-Extension-Fam with others in between.

PC15 separates TECH6: USUHS/PCR-free/ADSP-FUS2-Replicates sample from others.

```{r pc_plots_1415_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC14", "PC15", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC14
pc14_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC14_sd = sd(PC14),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC14_sd))

# Find replicate with largest spread on PC15
pc15_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC15_sd = sd(PC15),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC15_sd))

top_pc14_rep <- pc14_spreads$REP[1]
top_pc15_rep <- pc15_spreads$REP[1]

cat("**Largest spread on PC14:**", top_pc14_rep, "(SD =", sprintf("%.5f", pc14_spreads$PC14_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc14_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC14, PC15) %>%
  arrange(PC14)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC15:**", top_pc15_rep, "(SD =", sprintf("%.5f", pc15_spreads$PC15_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc15_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC14, PC15) %>%
  arrange(PC15)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

### D.6.6 PC15 vs PC16

PC16 separates TECH1: WASHU/PCR-free/ADSP-Extension-CC sample from others.

```{r pc_plots_1516_clusters, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC15", "PC16", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)

cat("\n\n**Replicate Groups with Largest Spreads:**\n\n")

# Find replicate with largest spread on PC15
pc15_spreads_1516 <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC15_sd = sd(PC15),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC15_sd))

# Find replicate with largest spread on PC16
pc16_spreads <- plot_data %>%
  group_by(REP) %>%
  summarise(
    N = n(),
    PC16_sd = sd(PC16),
    .groups = 'drop'
  ) %>%
  arrange(desc(PC16_sd))

top_pc15_rep_1516 <- pc15_spreads_1516$REP[1]
top_pc16_rep <- pc16_spreads$REP[1]

cat("**Largest spread on PC15:**", top_pc15_rep_1516, "(SD =", sprintf("%.5f", pc15_spreads_1516$PC15_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc15_rep_1516) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC15, PC16) %>%
  arrange(PC15)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))

cat("\n**Largest spread on PC16:**", top_pc16_rep, "(SD =", sprintf("%.5f", pc16_spreads$PC16_sd[1]), ")\n\n")
rep_samples <- plot_data %>%
  filter(REP == top_pc16_rep) %>%
  select(ID, REP, Cohort_Site, Study, Round, PCR.free, Sequencing_Center, PC15, PC16) %>%
  arrange(PC16)
print(knitr::kable(rep_samples, row.names = FALSE, digits = 5))
```

**Interpretation:** The eigenvector plots (D.1-D.4) visualize the first five PCs, where technical replicates cluster tightly together. These explain most of the variation as they should. TECH1, TECH2 and TECH3 are from the same family (mother and full-sibling offspring), which explains why they cluster together most tightly in the earliest PCs. Notably, PC1 splits the replicate groups into two sets: those with the PCR-amplified prep and those without.

The analysis of PC6-PC16 (D.6) in the replicate split region is where technical factors begin to separate samples. The consistent outlier in all replicate groups are samples from the ADSP-Extension-Fam study sequenced at WASHU with the PCR-free prep. The ADSP-Extension-CC study at WASHU with PCR-free prep also showed differences in multiple groups, as did, the LASI-DAD sample in TECH4 and the ADSP-FUS2-Replicates sample in TECH6. Prep does not explain any differences in the replicate-split region, perhaps because differences due to prep were captured by PC1.

# E. Summary

```{r summary_stats, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
library(dplyr)

# Calculate summary statistics
early_pcs_data <- spreads_long %>%
  filter(PC_num <= 5)

max_early_sd <- max(early_pcs_data$SD, na.rm = TRUE)

high_sd_reps <- early_pcs_data %>%
  filter(SD > sd_threshold_early_pcs) %>%
  distinct(Replicate)

n_high_early_sd <- nrow(high_sd_reps)

pc_at_max_eigenvalue <- which.max(eigenvalues)
first_pc_below_1 <- min(which(eigenvalues < 1), na.rm = TRUE)

cat("**Key Findings:**\n\n")
cat("- **Dataset**: ", n_samples, " technical replicate samples across ", n_groups, " groups\n")
cat("- **Principal Components**: ", num_pcs, " PCs calculated\n")
cat("- **Variance**: First 5 PCs explain ", round(sum(var_explained[1:5]), 2), "% of variance\n")
cat("- **Kaiser Criterion**: ", n_pcs_gt1, " PCs with eigenvalue > 1\n\n")

cat("**Technical Replicate Quality:**\n\n")
cat("- Very low within-group SD in early PCs (1-5): ", format(max_early_sd, scientific = TRUE, digits = 3), "\n")
cat("- Expected replicate split point: PC",n_groups - 1, "\n\n")

cat("**Technical Factors Examined:**\n\n")
cat("- PCR-free vs standard library preparation\n")
cat("- Sequencing round differences\n")
cat("- Sequencing center differences\n")
cat("- Cohort/Site effects\n\n")

if (n_high_early_sd == 0) {
  assessment <- "excellent"
} else if (n_high_early_sd / n_groups < 0.1) {
  assessment <- "very good"
} else if (n_high_early_sd / n_groups < 0.25) {
  assessment <- "good"
} else {
  assessment <- "variable"
}


cat("**Key Observations:**\n\n")
cat("1. **PCR Prep Effect on Early PCs**: PC1 cleanly separates replicate groups by library preparation method, dividing samples into PCR-amplified (TECH1, TECH2, TECH3) vs. PCR-free (TECH4, TECH5, TECH6) groups. This is the dominant source of variation in the dataset.\n\n")

cat("2. **Between-Group Separation**: PC2-PC5 separate individual replicate groups from each other, with technical replicates clustering tightly within their groups. This represents biological differences between individuals. Notably, TECH1, TECH2, and TECH3 are from teh same family (mother and full-sibling offspring), which explains their particularly tight clustering and is captured by PC4-PC5 which separate the mother from her offspring and the two siblings from each other.\n\n")

cat("3. **Replicate Split Pattern**: The expected replicate split occurs after PC",n_groups - 1, ", consistent with theoretical expectations if most variation is among the six replicate groups rather than with. Before this point, technical replicates cluster tightly together; after this point, within-group technical variation becomes observable.\n\n")

cat("4. **Study and Sequencing Center Effects in Split Region**: In the replicate split region (PC6-PC16), the consistent outliers are samples from specific Study/Sequencing_Center combinations:\n")
cat("   - **ADSP-Extension-Fam samples at WASHU** appear as outliers in multiple replicate groups\n")
cat("   - **ADSP-Extension-CC samples at WASHU** also show separation in several groups\n")
cat("   - **LASI-DAD sample in TECH4** (MEDGENOME sequencing center)\n")
cat("   - **ADSP-FUS2-Replicates sample in TECH6** (USUHS sequencing center)\n")
cat("   - Notably, **PCR prep does NOT explain differences in the replicate-split region** - the separation is driven by Study and Sequencing_Center combinations\n\n")

cat("5. **Quality Control Recommendations**:\n")
cat("   - The consistent outlier pattern for WASHU/ADSP-Extension-Fam samples suggests systematic differences that should be investigated\n")
cat("   - Consider whether samples from specific Study/Sequencing_Center combinations require batch effect correction or exclusion when using those datasets\n")
cat("   - Later sequencing rounds show excellent concordance in PC space.\n\n")

```
