---
title: "Analysis of PCA from Technical Replicates from ADSP R5"
author: "ERM"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 4
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# A. R5 Technical Replicate Data

## A.1 Overview

Technical replicates are samples from the same individual processed through independent library preparations and sequencing runs. This document analyzes PCA results from *technical replicates* in the ADSP R5 post-QC dataset to assess technical variation and sequencing quality.

**What this analysis does:**

1. **Data Loading**: Reads PCA eigenvectors/eigenvalues and covariate data
2. **Quality Assessment**: Examines variance distribution via scree plots
3. **Replicate Concordance**: Calculates within-group standard deviations across all PCs
4. **Technical Issues**: Identifies replicate groups with unexpectedly high variation
5. **Visualization**: Creates PC scatter plots colored by technical factors

**Key expectations for technical replicates:**

- **Near-zero variation** in early PCs (biological signal should be identical)
- **Replicate split** occurs around PC (n_groups - 1) where noise dominates
- **Technical factors** (PCR-free vs standard, sequencing round) may explain any observed variation

## A.2 Replicate PCA Protocol

We follow the same protocol as for non-technical replicates:

[**Replicate PCA Protocol**]{.underline}

A principal components analysis (PCA) was implemented using replicate samples identified through identity-by-descent analysis. Only replicate groups that include at least one sample from this sequencing round (r5) are included.

-   Select variants (SNVs or INDELs) meeting all of the following criteria:

    -   polymorphic among replicates

    -   no genotype missingness across replicates

    -   minor allele frequency (MAF)\>0.05

    -   QC criteria from fields in \*.summary file: `GLNexusPass`=1 and `CallBad`=0 and `HiDepth`=0

-   LD prune selected variants (r2 threshold of 0.8)

-   Conduct PCA (EIGENSOFT/EIGENSTRAT) to obtain all eigenvectors (PCs) and eigenvalues

## A.3 Data Description

### A.3.1 File Parameters

```{r file_parameters, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide"}
# ============================================================================
# FILE PATHS AND PARAMETERS - MODIFY AS NEEDED
# ============================================================================

# ---- INPUT FILES ----
# PCA file format: CSV with first row containing eigenvalues in columns 4+
#                  Subsequent rows contain: ID, Rep, index, PC1, PC2, ..., PCn
pca_file <- "data/TECH_pruned_evec.csv"

# Covariate file format: CSV with columns including:
#   Required: SampleID, Cohort_Site, Study, Round, Sex, Race, Ethnicity
#   Optional: DX_harmonized, Age_harmonized, APOE_reported, PCR.free, Technical_Replicate
covariate_file <- "data/r5_replicate_covariates_TechReps.csv"

# ---- OUTPUT FILES ----
processed_pca_output <- "processed_tech_pca_data.csv"  # Cleaned PCA data with PC column names
replicate_spreads_cache <- "tech_rep_spreads.rds"      # Cache of SD calculations (delete to recalculate)

# ---- ANALYSIS PARAMETERS ----
# PC Region Definitions - MODIFY THESE TO CHANGE ANALYSIS REGIONS
early_pcs_range <- c(1, 5)           # Early PCs region (PC1-PC5)
split_region_range <- c(6, 16)       # Replicate split region (PC6-PC16)
late_pcs_range <- c(17, NA)          # Late PCs region (PC17-end, NA = to last PC)

# Threshold for flagging high SD in early PCs (indicates technical issues)
# For technical replicates, expect SDs < 1e-4 in early PCs
sd_threshold_early_pcs <- 1e-4

# Threshold for flagging high SD in replicate split region
sd_threshold_split_region <- 0.1

# Percentile cutoff for highlighting top spreads in plots (0.90 = top 10%)
top_percentile_cutoff <- 0.90

# Number of PCs to show in zoomed scree plot
n_pcs_zoom <- 10

# ============================================================================

# Quick reference for threshold values:
# - sd_threshold_early_pcs: Increase if too many false positives (e.g., 1e-3)
# - sd_threshold_split_region: Decrease to be more stringent (e.g., 0.05)
# ============================================================================

# Validate input files exist
if (!file.exists(pca_file)) {
  stop("PCA file not found: ", pca_file)
}
if (!file.exists(covariate_file)) {
  stop("Covariate file not found: ", covariate_file)
}

cat("**Input files validated:**\n")
cat("- PCA file:", pca_file, "\n")
cat("- Covariate file:", covariate_file, "\n\n")
```

### A.3.2 Data Integrity Checks

```{r data_checks, message=FALSE, warning=FALSE, echo=FALSE}
# Pre-read checks
pca_data_raw <- read.csv(pca_file, header = FALSE, nrows = 5)

cat("**Data Format Checks:**\n\n")
cat("- PCA file columns:", ncol(pca_data_raw), "\n")
cat("- Expected: Column 1 = ID, Column 2 = Rep, Column 3 = index, Columns 4+ = eigenvalues/PCs\n")

# Check for minimum required columns
if (ncol(pca_data_raw) < 4) {
  stop("ERROR: PCA file has fewer than 4 columns. Expected: ID, Rep, index, eigenvalues/PCs")
}

cat("- Format appears correct\n\n")
```

### A.3.3 Replicate Sets

```{r get_pca_data, message=FALSE, warning=FALSE, echo=FALSE}
# Read PCA results file
df <- read.csv(
  pca_file,
  header = FALSE,
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Extract eigenvalues (row 1, col 4+)
eigenvalues <- as.numeric(df[1, 4:ncol(df)])
headers <- as.character(df[1, ])
ev_df <- df[-1, ]
colnames(ev_df) <- headers

# Rename EV columns to PC1..PCn
num_pcs <- ncol(ev_df) - 3
pc_names <- paste0("PC", seq_len(num_pcs))
colnames(ev_df)[4:ncol(ev_df)] <- pc_names

# Convert PCs to numeric
ev_df[pc_names] <- lapply(ev_df[pc_names], as.numeric)

# Save processed data
write.csv(ev_df, processed_pca_output, row.names = FALSE)

# Standardize column names for consistency
if ("Rep" %in% names(ev_df)) {
  names(ev_df)[names(ev_df) == "Rep"] <- "REP"
}
if ("ID" %in% names(ev_df)) {
  # ID column is already correct
} else {
  stop("Expected 'ID' column not found in PCA data")
}

# Verify we have the expected columns
expected_cols <- c("ID", "REP", "index")
missing_cols <- setdiff(expected_cols, names(ev_df))
if (length(missing_cols) > 0) {
  stop("Missing expected columns: ", paste(missing_cols, collapse = ", "))
}

# Count replicate groups
library(dplyr)
library(knitr)

n_groups <- length(unique(ev_df$REP))
n_samples <- nrow(ev_df)

# Create detailed replicate group table
rep_group_summary <- ev_df %>%
  group_by(REP) %>%
  summarise(
    N_Samples = n(),
    Sample_IDs = paste(ID, collapse = ", ")
  ) %>%
  arrange(REP)

# Add group type
rep_group_summary <- rep_group_summary %>%
  mutate(
    Group_Type = case_when(
      N_Samples == 2 ~ "Pair",
      N_Samples == 3 ~ "Trio",
      N_Samples == 4 ~ "Quad",
      TRUE ~ paste0("Group_", N_Samples)
    )
  ) %>%
  select(REP, Group_Type, N_Samples, Sample_IDs)

cat("**Technical Replicate Summary:**\n\n")
cat("- **", n_groups, "Replicate Groups**\n")
cat("- **", n_samples, "Total Samples**\n")
cat("- **", num_pcs, "Principal Components**\n\n")

# Print table
kable(rep_group_summary,
      col.names = c("Replicate", "Type", "N", "Sample IDs"),
      caption = "Replicate Group Details")

cat("\n**Expected Analysis Behavior:**\n\n")
cat("- Technical replicates should show near-zero variation in early PCs\n")
cat("- Replicate split expected around PC", n_groups - 1, "(n_groups - 1)\n")
cat("- SD > ", sd_threshold_early_pcs, " in early PCs indicates potential technical issues\n\n")
```

### A.3.4 Covariates

```{r get_cov_data, message=FALSE, warning=FALSE, echo=FALSE}
library(dplyr)

# Read covariate file and filter to samples in PCA
covariates_raw <- read.csv(covariate_file)

# Check for samples in PCA but not in covariates
samples_missing_cov <- setdiff(ev_df$ID, covariates_raw$SampleID)
if (length(samples_missing_cov) > 0) {
  warning("WARNING: ", length(samples_missing_cov), " samples in PCA are missing from covariate file")
  cat("Samples missing covariates:\n")
  print(head(samples_missing_cov, 10))
  if (length(samples_missing_cov) > 10) {
    cat("... and", length(samples_missing_cov) - 10, "more\n")
  }
}

covariates_filtered <- covariates_raw %>%
  filter(SampleID %in% ev_df$ID) %>%
  select(SampleID, Cohort_Site, Study, Round, Sex, DX_harmonized, Age_harmonized,
         Age_baseline, APOE_reported, Race, Ethnicity, Technical_Replicate, PCR.free)

# Recode Ethnicity and Race for readability
covariates_filtered <- covariates_filtered %>%
  mutate(
    Ethnicity = recode(Ethnicity, `0` = "NH", `1` = "H"),
    Race = recode(Race, `1` = "AI", `2` = "As", `4` = "Bl", `5` = "W", `6` = "Other")
  )

cat("**Covariate Summary:**\n\n")
cat("- Samples with covariate data:", nrow(covariates_filtered), "\n")
cat("- Cohort/Sites represented:", length(unique(covariates_filtered$Cohort_Site)), "\n")
cat("- Studies represented:", length(unique(covariates_filtered$Study)), "\n")
cat("- Sequencing rounds:", paste(sort(unique(covariates_filtered$Round)), collapse = ", "), "\n")
if("PCR.free" %in% names(covariates_filtered)) {
  pcr_free_summary <- table(covariates_filtered$PCR.free, useNA = "ifany")
  cat("- Library prep types:\n")
  for (type in names(pcr_free_summary)) {
    if (!is.na(type)) {
      cat("    - ", type, ": ", pcr_free_summary[type], "\n", sep = "")
    } else {
      cat("    - Missing: ", pcr_free_summary[type], "\n", sep = "")
    }
  }
}
cat("\n")
```

# B. Scree Plots

For technical replicates, we expect minimal variance overall since these are the same individuals. Most variance should reflect library prep and sequencing differences, not biological variation.

```{r scree_plot, message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)

scree_data <- data.frame(
  Component = 1:length(eigenvalues),
  Eigenvalue = eigenvalues
)

var_explained <- (eigenvalues / sum(eigenvalues)) * 100
n_pcs_gt1 <- sum(eigenvalues > 1)

cat("- First 3 PCs explain", round(sum(var_explained[1:3]), 2), "% of variance\n")
cat("- Number of PCs with eigenvalue > 1 (Kaiser criterion):", n_pcs_gt1, "\n")
cat("- Expected replicate split at PC", n_groups - 1, "\n\n")

# Full scree plot
ggplot(scree_data, aes(x = Component, y = Eigenvalue)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "orange") +
  geom_vline(xintercept = n_samples - 1, linetype = "dashed", color = "red") +
  geom_vline(xintercept = n_groups - 1, linetype = "dashed", color = "blue") +
  annotate("text", x = n_samples - 1, y = max(scree_data$Eigenvalue) * 0.85, 
           label = paste(n_samples - 1, "= (Samples - 1)"), color = "red", angle = 90, vjust = -0.5) +
  annotate("text", x = n_groups - 1, y = max(scree_data$Eigenvalue) * 0.85, 
           label = paste(n_groups - 1, "= (Groups - 1)"), color = "blue", angle = 90, vjust = -0.5) +
  labs(title = "Scree Plot - Technical Replicates", x = "PC Number", y = "Eigenvalue") +
  theme_minimal()

# Zoom PC1-n
ggplot(scree_data, aes(x = Component, y = Eigenvalue)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "orange") +
  coord_cartesian(xlim = c(1, min(n_pcs_zoom, length(eigenvalues)))) +
  labs(title = paste0("Scree Plot (Zoomed PC1-PC", min(n_pcs_zoom, length(eigenvalues)), ")"),
       x = "PC Number", y = "Eigenvalue") +
  theme_minimal()
```

# C. Replicate Splits

For technical replicates, within-group SDs should be extremely close to zero across all PCs until the replicate split point, as these represent the same biological sample.

```{r setup_replicate_functions, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide", results='asis'}
library(tidyr)
library(dplyr)
library(purrr)
library(kableExtra)
library(tibble)
library(RColorBrewer)

calculate_replicate_spread <- function(data_df, rep_col, pc_cols, cache_file = NULL) {

  if (!is.null(cache_file) && file.exists(cache_file)) {
    message("Loading cached replicate spreads from ", cache_file)
    return(readRDS(cache_file))
  }

  # Validate inputs
  if (!rep_col %in% names(data_df)) {
    stop("Column '", rep_col, "' not found in data. Available columns: ", paste(names(data_df), collapse = ", "))
  }

  spread_list <- list()
  unique_reps <- unique(data_df[[rep_col]])

  for (rep in unique_reps) {
    rep_data <- data_df[data_df[[rep_col]] == rep, ]

    if (nrow(rep_data) < 2) next

    # Classify replicate type based on group size
    n_samples <- nrow(rep_data)
    rep_type <- if (n_samples == 2) {
      "Pair"
    } else if (n_samples == 3) {
      "Trio"
    } else if (n_samples == 4) {
      "Quad"
    } else {
      paste0("Group_", n_samples)
    }

    spreads <- data.frame(
      Replicate = rep,
      N_samples = n_samples,
      Rep_Type = rep_type
    )

    for (pc in pc_cols) {
      pc_values <- rep_data[[pc]]
      spreads[[paste0(pc, "_sd")]] <- sd(pc_values)
    }

    spread_list[[length(spread_list) + 1]] <- spreads
  }

  # Check if we found any replicate groups
  if (length(spread_list) == 0) {
    stop("No replicate groups with 2+ samples found. Check that '", rep_col, "' column contains replicate group IDs.")
  }

  all_spreads <- do.call(rbind, spread_list)

  if (!is.null(cache_file)) {
    message("Saving replicate spreads to ", cache_file)
    saveRDS(all_spreads, cache_file)
  }

  return(all_spreads)
}

plot_pc_range <- function(data_df, pc_min, pc_max, top_flag = NULL, sd_threshold = 1e-4) {

  data_sub <- data_df %>%
    filter(PC_num >= pc_min, PC_num <= pc_max)

  pc_levels <- data_sub %>%
    distinct(PC, PC_num) %>%
    arrange(PC_num) %>%
    pull(PC)

  data_sub <- data_sub %>%
    mutate(PC = factor(PC, levels = pc_levels))

  # Get unique replicates and assign colors
  unique_reps <- sort(unique(data_sub$Replicate))
  n_reps <- length(unique_reps)

  # Use a color palette that can handle many categories
  if (n_reps <= 8) {
    rep_colors <- RColorBrewer::brewer.pal(max(3, n_reps), "Set2")[1:n_reps]
  } else if (n_reps <= 12) {
    rep_colors <- RColorBrewer::brewer.pal(n_reps, "Set3")
  } else {
    rep_colors <- rainbow(n_reps)
  }
  names(rep_colors) <- unique_reps

  p <- ggplot(data_sub, aes(x = PC, y = SD))

  if (!is.null(top_flag)) {
    p <- p + geom_point(
      data = data_sub %>% filter(!.data[[top_flag]]),
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.4,
      size = 1.2
    )
    p <- p + geom_point(
      data = data_sub %>% filter(.data[[top_flag]]),
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.8,
      size = 1.5
    )
  } else {
    p <- p + geom_point(
      data = data_sub,
      aes(color = Replicate),
      position = position_dodge(width = 0.5),
      alpha = 0.6,
      size = 1.2
    )
  }

  theme_settings <- if ((pc_max - pc_min) > 50) {
    theme(axis.text.x = element_blank())
  } else {
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
  }

  p <- p +
    labs(
      title = paste0("Technical Replicate SDs: PC", pc_min, "–PC", pc_max),
      x = "Principal Component",
      y = "Standard Deviation",
      color = "Replicate"
    ) +
    scale_color_manual(values = rep_colors) +
    scale_y_continuous(limits = c(0, NA)) +
    theme_minimal() +
    theme_settings

  return(p)
}

print_high_sd_reps <- function(data_df, plot_data, pc_min, pc_max, sd_threshold = 0.1) {
  
  data_sub <- data_df %>%
    filter(PC_num >= pc_min, PC_num <= pc_max, SD > sd_threshold) %>%
    arrange(desc(SD))
  
  if (nrow(data_sub) == 0) {
    cat("No replicates with SD >", sd_threshold, "in PC", pc_min, "-", pc_max, "\n\n")
    return(invisible(NULL))
  }
  
  joined_data <- data_sub %>%
    left_join(plot_data, by = c("Replicate" = "REP"))
  
  all_pc_colnames <- names(joined_data) %>% stringr::str_subset("^PC\\d+$")
  
  enriched <- joined_data %>%
    mutate(
      Values = purrr::map2(
        .x = seq_len(nrow(joined_data)),
        .y = PC_num,
        .f = function(row_index, pc_num_value) {
          target_col_name <- paste0("PC", pc_num_value)
          return(joined_data[[target_col_name]][row_index])
        }
      )
    )
  
  summary_table <- enriched %>%
    arrange(PC_num, SD) %>%
    transmute(
      PC,
      ID,
      Replicate,
      Cohort_Site,
      Round,
      PCR.free,
      SD,
      Values = sapply(Values, function(x) paste(x, collapse = ", "))
    )
  
  cat("\n**Technical Replicates with SD >", sd_threshold, "for PC", pc_min, "-", pc_max, ":**\n\n")
  print(knitr::kable(summary_table, row.names = FALSE, digits = 5))
  
  return(invisible(summary_table))
}

find_variable_replicates <- function(replicate_spreads, threshold = 0) {
  
  sd_cols <- grep("_sd$", names(replicate_spreads), value = TRUE)
  
  variable_reps <- replicate_spreads %>%
    mutate(max_sd = pmax(!!!syms(sd_cols))) %>%
    filter(max_sd > threshold) %>%
    select(Replicate, N_samples, max_sd) %>%
    arrange(desc(max_sd))
  
  cat("\n**Replicate groups with SD >", threshold, ":**\n\n")
  cat("Total:", nrow(variable_reps), "out of", nrow(replicate_spreads), "groups\n\n")
  
  print(knitr::kable(variable_reps, row.names = FALSE, digits = 4))
  
  return(invisible(variable_reps))
}

# Prepare plot_data
collapse_rare_levels <- function(column, threshold = 75) {
  level_counts <- table(column)
  rare_levels <- names(level_counts[level_counts <= threshold])
  
  collapsed <- as.character(column)
  collapsed[collapsed %in% rare_levels] <- "Other"
  collapsed <- as.factor(collapsed)
  
  return(collapsed)
}

covariates_filtered$Cohort_Site_collapsed <- collapse_rare_levels(covariates_filtered$Cohort_Site)
covariates_filtered$Study_collapsed <- collapse_rare_levels(covariates_filtered$Study)

plot_data <- merge(ev_df, covariates_filtered, by.x = "ID", by.y = "SampleID")

plot_data <- plot_data %>%
  mutate(across(c(Cohort_Site, Study, Round, Race, Ethnicity, Sex, DX_harmonized,
                  Cohort_Site_collapsed, Study_collapsed, PCR.free), as.factor))

plot_data$Ethnicity <- factor(plot_data$Ethnicity, levels = c("NH", "H", "Other"))
plot_data$Race <- factor(plot_data$Race, levels = c("W", "Bl", "As", "AI", "Other"))

# Ensure PCR.free has proper factor levels
if("PCR.free" %in% names(plot_data)) {
  plot_data$PCR.free <- factor(plot_data$PCR.free,
                                levels = c("PCRFree", "PCRAmplified"),
                                labels = c("PCR-free", "PCR-amplified"))
}
```

## C.1 Overall Visualization

```{r rep_splits_overall, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}

pc_cols <- paste0("PC", 1:num_pcs) 

replicate_spreads <- calculate_replicate_spread(
  plot_data,
  rep_col = "REP",
  pc_cols = pc_cols,
  cache_file = replicate_spreads_cache
)

spreads_long <- replicate_spreads %>%
  select(Replicate, N_samples, ends_with("_sd")) %>%
  pivot_longer(cols = ends_with("_sd"),
               names_to = "PC",
               values_to = "SD") %>%
  mutate(PC = gsub("_sd", "", PC),
         PC_num = as.numeric(gsub("PC", "", PC)))

spreads_long <- spreads_long %>%
  group_by(PC) %>%
  mutate(pcut = quantile(SD, top_percentile_cutoff, na.rm = TRUE),
         Top = SD > pcut) %>%
  ungroup()

plot_pc_range(spreads_long, pc_min = 1, pc_max = num_pcs)

variable_reps <- find_variable_replicates(replicate_spreads, threshold = 0)
```

## C.2 Early PCs (Technical Quality)

For technical replicates, early PCs should show near-zero within-group variation. Any elevation suggests technical issues.

```{r rep_splits_early, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
plot_pc_range(spreads_long, pc_min = early_pcs_range[1], pc_max = early_pcs_range[2])
print_high_sd_reps(spreads_long, plot_data, pc_min = early_pcs_range[1], pc_max = early_pcs_range[2],
                   sd_threshold = sd_threshold_early_pcs)
```

## C.3 Replicate Split Region

```{r rep_splits_middle, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
plot_pc_range(spreads_long, pc_min = split_region_range[1], pc_max = split_region_range[2])
print_high_sd_reps(spreads_long, plot_data, pc_min = split_region_range[1], pc_max = split_region_range[2],
                   sd_threshold = sd_threshold_split_region)
```

## C.4 Late PCs

```{r rep_splits_late, echo=FALSE, message=FALSE, warning=FALSE, eval=TRUE, results='asis'}
late_pc_end <- if (is.na(late_pcs_range[2])) num_pcs else late_pcs_range[2]
plot_pc_range(spreads_long, pc_min = late_pcs_range[1], pc_max = late_pc_end)
print_high_sd_reps(spreads_long, plot_data, pc_min = late_pcs_range[1], pc_max = late_pc_end,
                   sd_threshold = sd_threshold_split_region)
```

# D. Eigenvector Plots

```{r setup_plotting_functions, message=FALSE, warning=FALSE, echo=FALSE, class.source="fold-hide"}
library(ggplot2)
library(kableExtra)

plot_pcs <- function(df, pc_x, pc_y, color_by, shape_by = NULL,
                     color_title = NULL, shape_title = NULL,
                     var_explained = NULL) {

  if (is.null(color_title)) color_title <- color_by
  if (is.null(shape_title)) shape_title <- shape_by

  x_label <- pc_x
  y_label <- pc_y
  if (!is.null(var_explained)) {
    pc_x_num <- as.numeric(gsub("PC", "", pc_x))
    pc_y_num <- as.numeric(gsub("PC", "", pc_y))
    x_label <- paste0(pc_x, " (", round(var_explained[pc_x_num], 2), "%)")
    y_label <- paste0(pc_y, " (", round(var_explained[pc_y_num], 2), "%)")
  }

  p <- ggplot(df, aes(x = .data[[pc_x]], y = .data[[pc_y]],
                      color = .data[[color_by]]))

  if (!is.null(shape_by)) {
    p <- p + geom_point(aes(shape = .data[[shape_by]]), size = 1.5, stroke = 0.5)
    p <- p + scale_shape_manual(values = c(1, 2, 0, 5, 6), na.value = 4)
  } else {
    p <- p + geom_point(size = 1.5, stroke = 0.5, shape = 1)
  }

  # Use custom colors for REP column (replicate groups)
  if (color_by == "REP") {
    unique_reps <- sort(unique(df[[color_by]]))
    n_reps <- length(unique_reps)

    if (n_reps <= 8) {
      rep_colors <- RColorBrewer::brewer.pal(max(3, n_reps), "Set2")[1:n_reps]
    } else if (n_reps <= 12) {
      rep_colors <- RColorBrewer::brewer.pal(n_reps, "Set3")
    } else {
      rep_colors <- rainbow(n_reps)
    }
    names(rep_colors) <- unique_reps

    p <- p + scale_color_manual(values = rep_colors)
  }

  p <- p +
    labs(x = x_label, y = y_label, color = color_title, shape = shape_title) +
    theme_minimal()

  return(p)
}
```

## D.1 PC1 vs PC2

Technical replicates should cluster tightly. Any separation could indicate technical issues (library prep, PCR-free vs standard).

### D.1.1 By Technical Factors

```{r pc_plots_12_tech, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC1", "PC2", color_by = "Round", var_explained = var_explained)
plot_pcs(plot_data, "PC1", "PC2", color_by = "PCR.free", var_explained = var_explained)
```

### D.1.2 By Replicate Group

```{r pc_plots_12_rep, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC1", "PC2", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

### D.1.3 By Cohort/Site

```{r pc_plots_12_cohort, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC1", "PC2", color_by = "Cohort_Site", var_explained = var_explained)
```

## D.2 PC2 vs PC3

### D.2.1 By Technical Factors

```{r pc_plots_23_tech, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC2", "PC3", color_by = "Round", var_explained = var_explained)
plot_pcs(plot_data, "PC2", "PC3", color_by = "PCR.free", var_explained = var_explained)
```

### D.2.2 By Replicate Group

```{r pc_plots_23_rep, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC2", "PC3", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.3 PC3 vs PC4

### D.3.1 By Technical Factors

```{r pc_plots_34_tech, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC3", "PC4", color_by = "Round", var_explained = var_explained)
plot_pcs(plot_data, "PC3", "PC4", color_by = "PCR.free", var_explained = var_explained)
```

### D.3.2 By Replicate Group

```{r pc_plots_34_rep, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
plot_pcs(plot_data, "PC3", "PC4", color_by = "REP",
         color_title = "Replicate", var_explained = var_explained)
```

## D.4 Additional Covariates

```{r pc_plots_additional, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
# Plot by study if there are multiple studies
if (length(unique(plot_data$Study)) > 1) {
  plot_pcs(plot_data, "PC1", "PC2", color_by = "Study_collapsed", var_explained = var_explained)
}

# Plot by sex if variable
if (length(unique(plot_data$Sex)) > 1) {
  plot_pcs(plot_data, "PC1", "PC2", color_by = "Sex", var_explained = var_explained)
}
```

# E. Summary

```{r summary_stats, message=FALSE, warning=FALSE, echo=FALSE, results='asis'}
library(dplyr)

# Calculate summary statistics
early_pcs_data <- spreads_long %>%
  filter(PC_num <= 40)

max_early_sd <- max(early_pcs_data$SD, na.rm = TRUE)

high_sd_reps <- early_pcs_data %>%
  filter(SD > sd_threshold_early_pcs) %>%
  distinct(Replicate)

n_high_early_sd <- nrow(high_sd_reps)

pc_at_max_eigenvalue <- which.max(eigenvalues)
first_pc_below_1 <- min(which(eigenvalues < 1), na.rm = TRUE)

cat("**Key Findings:**\n\n")
cat("- **Dataset**: ", n_samples, " technical replicate samples across ", n_groups, " groups\n")
cat("- **Principal Components**: ", num_pcs, " PCs calculated\n")
cat("- **Variance**: First 3 PCs explain ", round(sum(var_explained[1:3]), 2), "% of variance\n")
cat("- **Kaiser Criterion**: ", n_pcs_gt1, " PCs with eigenvalue > 1\n\n")

cat("**Technical Replicate Quality:**\n\n")
cat("- Maximum SD in early PCs (1-40): ", format(max_early_sd, scientific = TRUE, digits = 3), "\n")
cat("- Replicate groups with SD > ", sd_threshold_early_pcs, " in early PCs: ", n_high_early_sd, " out of ", n_groups, "\n")
cat("- Expected replicate split point: PC", n_groups - 1, "\n\n")

cat("**Technical Factors to Examine:**\n\n")
cat("- PCR-free vs standard library preparation\n")
cat("- Sequencing round differences\n")
cat("- Cohort/Site effects\n\n")

if (n_high_early_sd == 0) {
  assessment <- "excellent"
} else if (n_high_early_sd / n_groups < 0.1) {
  assessment <- "very good"
} else if (n_high_early_sd / n_groups < 0.25) {
  assessment <- "good"
} else {
  assessment <- "variable"
}

cat("**Overall Assessment**: Technical replicates demonstrate **", assessment, "** concordance.\n")
if (n_high_early_sd > 0) {
  cat("Review the flagged replicate groups in Section C.2 for potential technical issues.\n")
}
```
